Palmer amaranth (*Amaranthus palmeri*) adaptation to agroecossystems

### 2021-08-13 Manuscript/Manuscript.Rmd
as

Light and temperature are likely the main environment requirements for Palmer amaranth successful grow and development. Palmer amaranth is reported with an extended germination period [@jha2010]. Germination of Palmer amaranth is triggered by 18 C soil temperature [@keeley1987], and optimal germination and biomass production occur at 35/30 C day and night temperatures [@guo2003]. Palmer amaranth germination is also human-mediated by tillage timings and preemergence-applied herbicides [@chahal2021]. Water has not shown to limit Palmer amaranth fitness. Under continuous water stress, Palmer amaranth survived and produced at least 14000 seeds plant-1 [@chahal2018]. Seeds from Palmer amaranth growing with limited water conditions were heavier, less dormant, and prompt for germination [@matzrafi2021]. Also, growing conditions and management can impact sex dimorphism and flowering pattern in Palmer amaranth [@korres2017; @rumpa2019]. The continuous global temperature warming can impact agriculture and promote niches for Palmer amaranth invasion/adaptation into new environments. Currently, it is estimated that the greatest climatic risk of Palmer amaranth establishment are agronomic crops in Australia and Sub-Sahara Africa [@kistner2018]. Temperature is a key factor limiting Palmer amaranth expansion to cooler geographies [@briscoerunquist2019]; however, under future climate change Palmer amaranth is likely to expand northward into Canada and Northern Europe [@kistner2018; @briscoerunquist2019].

Palmer amaranth is already found in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the US, Palmer amaranth is well established at cotton fields [@garetson2019] and non-crop areas [@bagavathiannan2016] in the southern United States but its range is expanding northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a],  Michigan [@kohrt2017], Connecticut [@aulakh2021], and potential damaged to soybean fields in Illinois is reported [@davis2015]. In Iowa, a study showed that Palmer amaranth is not well adapted compared to waterhemp (*Amaranthus tuberculatus*) [@baker2021]. Successful cases of Palmer amaranth invasion and near to eradication is documented in Minnesota [@yu2021]. No Palmer amaranth actively growing was found in Canada; however, Palmer amaranth seeds was detected in sweet potato slips [@page2021]. Nonetheless, it seems fated the need to manage Palmer amaranth in agronomic crops throughout multiple environments in the near future. Strategies on Palmer amaranth management should encompass the agroecosystem level but not only attempts to eradicate the weed. Most tactics to manage Palmer amaranth are based on technology fixes [@scott2011], which are short-term (e.g., herbicide and/or tillage) rather than long-term weed management. Palmer amaranth management should be built on minimizing the species ability to adapt, grow and develop into agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to have higher impact on cotton yields than late established plants [@macrae2013]. In the northern states, Palmer amaranth impact on the agroecossystem is recent. Studies investigating Palmer amaranth in those locations is limited due to the plant classification as noxious weed species [@yu2021]. Nonetheless, the continuous Palmer amaranth dispersal and potential establishment across northern United States is concerning and warrant investigations on species morphology in such environments. Understanding Palmer amaranth biology and growing strategies under different agroecossystems can enhance our knowledge on species adaptation. It can also aid on designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and design resilient and sustainable farming systems [@maclaren2020]. Therefore, the objective of this study was to investigate the flowering pattern, biomass production, and height of Palmer amaranth growing under corn, soybean and fallow at two timings across five locations in the United States Midwest.

@@

Palmer amaranth height was more uniform across cohort timings, except when growing in corn (Figure 3B). Palmer amaranth achieve achieve 69.2 cm tall when growing at first cohort timing in bareground, which was not different to 70.7 cm tall at second cohort timing (P = 0.74). In addition, no difference in Palmer amaranth height (69.3 cm) was found in soybean at first cohort compared to fallow (P > 0.75). At second cohort in soybean, Palmer amaranth was near to 10 cm lower compared to its first cohort time (P = 0.04). The tallest and smallest Palmer amaranth plants were found in corn. Palmer amaranth reached 85.2 cm tall at first cohort and 38.2 cm at second cohort timings.

@@

Our study showed that Palmer amaranth biomass, height, flowering pattern and gender varied within agroecossystems and cohort timings. In general, Palmer amaranth produced more biomass and taller plants when growing at first cohort rather than second cohort. At first cohort, resources (e.g., soil nutrients) and conditions (e.g., light) were more timely available for both species, crop and weed. High biomass and taller Palmer amaranth plants are likely a weed strategy to compete for light in between crop rows in absence of canopy. In such condition, Palmer amaranth showed an extraordinary plasticity to adapt upon the agroecosystem. This is evident when comparing Palmer amaranth canopy shape, and its extended flowering pattern when growing into corn compared to soybean. The Palmer amaranth competition strategy was to mimic the crop grow and development (Figure 4). These results suggests that Palmer amaranth can fast evolve life-history traits to adapt  into agroecossystems and cultural practices, including nitrogen fertilization [@bravo2018]. Our results highlight the Palmer amaranth threat to agroecossystems as breeding more competitive crop varieties is likely to select more competitive biotypes [@bravo2017].

@@

The mechanisms of sex-determination in plant species is intriguing and arouse the curiosity of many scientists, including Darwin [@darwin1888]. In our study, the gender model performance was decent considering the biology of plant flowering. A 1:1 male and female sex ratio is a general evolutionary stable strategy for plant species perpetuation [@fisher1930]. However, a slight deviation from 1:1 sex ratio might occur in some dioecious species. For example, the dioecious *Halophila stipulacea* is a female-biased plant in its native habitat, but the naturalized *H. stipulacea* have a 1:1 ratio [@nguyen2018]. Naturalized of *H. stipulacea* reduced female-male ratio to expand into its non-native habitat [@nguyen2018]. Also, biotic and/or abiotic stress can influence plant sex determination. Palmer amaranth  male-to-female ratio was greater under high plant densities [@korres2017a] and after herbicide application [@rumpa2019]. Our model estimated that late flowering, heavier and taller Palmer amaranth plants deviated from 1:1 ratio in favor to female plants. It was reported that female Palmer amaranth plants invested more in height, stem and biomass while male invested more in leaf area and leaf dry weight under nutrient deficiency [@korres2017]. Our model also estimated more female plants in soybean and bareground compared to corn, which might be a result of plant biomass in each agroecossystem. Our results showed the influence of life-history and ecological traits on sexual dimorphism in Palmer amaranth. Sexual dimorphism is documented in other dioecious species [@barrett2013]. For example, stronger female plant competition and greater male tolerance to herbivory was reported in *Spinacia oleracea* [@perez-llorca2019]. Research on candidate genes for sex determination in *Amaranthus* species are currently underway but it is far to complete [@montgomery2021; @montgomery2019]. Further studies are also needed to understand the ecological basis of Palmer amaranth flowering, including the plant behavior under climate change.

>>>>

Light and temperature are likely the main environment requirements for Palmer amaranth successful grow and development. Palmer amaranth is reported with an extended germination period [@jha2010]. Germination of Palmer amaranth is triggered by 18 C soil temperature [@keeley1987], and optimal germination and biomass production occur at 35/30 C day and night temperatures [@guo2003]. Palmer amaranth germination is also human-mediated by tillage timings and preemergence-applied herbicides [@chahal2021]. Water has not shown to limit Palmer amaranth fitness. Under continuous water stress, Palmer amaranth survived and produced at least 14000 seeds plant-1 [@chahal2018]. Seeds from Palmer amaranth growing with limited water conditions were heavier, less dormant, and prompt for germination [@matzrafi2021]. Growing conditions and management practices also influence Palmer amaranth sex dimorphism and flowering pattern [@korres2017; @rumpa2019]. The continuous global temperature warming can impact agriculture and promote niches for Palmer amaranth invasion/adaptation into new environments. Currently, it is estimated that the greatest climatic risk of Palmer amaranth establishment are agronomic crops in Australia and Sub-Sahara Africa [@kistner2018]. Temperature is a key factor limiting Palmer amaranth expansion to cooler geographies [@briscoerunquist2019]; however, under future climate change Palmer amaranth is likely to expand northward into Canada and Northern Europe [@kistner2018; @briscoerunquist2019].

Palmer amaranth is already found in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the US, Palmer amaranth is well established at cotton fields [@garetson2019] and non-crop areas [@bagavathiannan2016] in the southern United States but its range is expanding northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a],  Michigan [@kohrt2017], Connecticut [@aulakh2021], and potential damaged to soybean fields in Illinois is reported [@davis2015]. In Iowa, a study showed that Palmer amaranth is still not well adapted compared to waterhemp (*Amaranthus tuberculatus*) [@baker2021]. Successful cases of Palmer amaranth invasion and near to eradication is documented in Minnesota [@yu2021]. No Palmer amaranth actively growing was found in Canada; however, Palmer amaranth seeds was detected in sweet potato slips [@page2021]. Nonetheless, it seems fated the need to manage Palmer amaranth in agronomic crops throughout multiple environments in the near future. Strategies on Palmer amaranth management should encompass the agroecosystem level but not only attempts to eradicate the weed. Most tactics to manage Palmer amaranth are based on technology fixes [@scott2011], which are short-term (e.g., herbicide and/or tillage) rather than long-term weed management. Palmer amaranth management should be built on minimizing the species ability to adapt, grow and develop into agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to have higher impact on cotton yields than late established plants [@macrae2013]. In the northern states, Palmer amaranth impact on the agroecossystem is recent. Studies investigating Palmer amaranth in those locations is limited due to the plant classification as noxious weed species [@yu2021]. Nonetheless, the continuous Palmer amaranth dispersal and potential establishment across northern United States is concerning and warrant investigations on species morphology in such environments. Understanding Palmer amaranth biology and growing strategies under different agroecossystems can enhance our knowledge on species adaptation. It can also aid on designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and design resilient and sustainable farming systems [@maclaren2020]. Therefore, the objective of this study was to investigate the flowering pattern, gender, biomass production, and height of Palmer amaranth growing under corn, soybean and fallow at two timings across five locations in the United States Midwest.

@@

Palmer amaranth height was more uniform across cohort timings, except when growing in corn (Figure 3B). Palmer amaranth achieved 69.2 cm tall when growing at first cohort timing in bareground, which was not different to 70.7 cm tall at second cohort timing (P = 0.74). In addition, no difference in Palmer amaranth height (69.3 cm) was found in soybean at first cohort compared to fallow (P > 0.75). At second cohort in soybean, Palmer amaranth was near to 10 cm lower compared to its first cohort time (P = 0.04). The tallest and smallest Palmer amaranth plants were found in corn. Palmer amaranth reached 85.2 cm tall at first cohort and 38.2 cm at second cohort timings.

@@

Our study showed that Palmer amaranth biomass, height, flowering pattern and gender varied within agroecossystems and cohort timings. In general, Palmer amaranth produced more biomass and taller plants when growing at first cohort rather than second cohort. At first cohort, resources (e.g., soil nutrients) and conditions (e.g., light) were more timely available for both species, crop and weed. High biomass and taller Palmer amaranth plants are likely a weed strategy to compete for light in between crop rows in absence of canopy. In such condition, Palmer amaranth showed an extraordinary plasticity to adapt upon the agroecosystem. This is evident when comparing Palmer amaranth canopy shape, and its extended flowering pattern when growing into corn compared to soybean. The Palmer amaranth competition strategy was to mimic the crop grow and development (Figure 4). These results suggests that Palmer amaranth can fast evolve life-history traits to adapt  into agroecossystems and cultural practices, including nitrogen fertilization [@bravo2018]. Our results highlight the Palmer amaranth as a threat to agroecossystems as breeding more competitive crop varieties is likely to select more competitive biotypes [@bravo2017].

@@

The mechanisms of sex-determination in plant species is intriguing and arouse the curiosity of many scientists, including Darwin [@darwin1888]. In our study, the gender model performance was decent considering the biology of plant flowering. A 1:1 male and female sex ratio is a general evolutionary stable strategy for plant species perpetuation [@fisher1930]. However, a slight deviation from 1:1 sex ratio might occur in some dioecious species. For example, the dioecious *Halophila stipulacea* is a female-biased plant in its native habitat, but the naturalized *H. stipulacea* have a 1:1 ratio [@nguyen2018]. Naturalized of *H. stipulacea* reduced female-male ratio to expand into its non-native habitat [@nguyen2018]. Also, biotic and/or abiotic stress can influence plant sex determination. Palmer amaranth  male-to-female ratio was greater under high plant densities [@korres2017a] and after herbicide application [@rumpa2019]. Our model estimated that late flowering, heavier and taller Palmer amaranth plants deviated from 1:1 ratio in favor to female plants. It was reported that female Palmer amaranth plants invested more in height, stem and biomass while male invested more in leaf area and leaf dry weight under nutrient deficiency [@korres2017]. Our model also estimated more female plants in soybean and bareground compared to corn, which might linked to plant competition strategy in each agroecossystem. Our results showed the influence of life-history and ecological traits on sexual dimorphism in Palmer amaranth. Sexual dimorphism is documented in other dioecious species [@barrett2013]. For example, stronger female plant competition and greater male tolerance to herbivory was reported in *Spinacia oleracea* [@perez-llorca2019]. Research on candidate genes for sex determination in *Amaranthus* species are currently underway but it is far to complete [@montgomery2021; @montgomery2019]. Further studies are also needed to understand the ecological basis of Palmer amaranth flowering, including the plant behavior under climate change.
